Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-10-01 papers/paper1/design.tex
updates through section 4

All user automata inherit from a common Base Automaton class that provides facilities for asynchronously creating and destroying child automata and bindings.
% The services provided by that common base class are indicated by the Base Automaton shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the system automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a user automaton.
Instead, an Automaton Manager provides a common higher-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

>>>>

All user automata inherit from a common base class (indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}) that provides facilities for asynchronously creating and destroying child automata and bindings.
The Base Automaton layer provides statically composed system actions that are used to communicate with the system automaton.
In most situations, the services provided by the Base Automaton layer are too low-level to be used directly by a user automaton.
Instead, an Automaton Manager provides a common higher-level interface for asynchronously managing a single child automaton using the services of the Base Automaton layer.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

### 2011-10-01 papers/paper1/programming_model.tex
updates through section 4

\section{I/O Automata Programming Model\label{programming_model}}

In this section, we show how I/O automata can be used to construct programs using ioa++.
Section~\ref{representation} describes how I/O automata are represented in the C++ language.
We then discuss the mechanisms available for dynamic composition in Section~\ref{dynamic_composition}.
Section~\ref{constellations} concludes with techniques for managing dynamic constellations of automata.

@@

All automata inherit from the common base class ioa::automaton which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept.
The metaprogramming facilities of the C++ language aid in the development of general purpose reusable components.
\fi

\paragraph*{Declaring and initializing state variables}
The state variables of an automaton are expressed as member variables of its corresponding class or class template.
\ifjournal
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Member variables are declared private, which is in keeping with the notion that the state of each automaton is independent, and are initialized using a constructor.
The constructor can also bootstrap the scheduler using the ioa::schedule function discussed below.
\ifjournal
For the asynch\_lcr\_automaton, the flags are both set initially false and the UID is initialized and added to the send queue.
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the ioa::schedule function discussed below.
\fi

\paragraph*{Declaring/defining actions}
Actions are member variables that model the \emph{action} concept.
Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
These traits are used to guide the execution of the methods by the model and to supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

@@

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
Another common pattern is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
\ifjournal
For the asynch\_lcr\_automaton, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this common member function.
\fi

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

@@

In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the ``nuts and bolts'' presented in the previous section.
This list is not exhaustive but rather serves as a distillation of our experiences so far working with I/O automata in ioa++.

\paragraph*{System action events}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can observe ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

\paragraph*{Error reporting}
The event semantics of I/O automata provide a natural mechanism for reporting errors.
An automaton that encounters an error can use an output action to signal the error.
The automaton can continue processing if the error is transient or stop if the error is permanent.
In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
Using this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.

@@

If all of the bindings of a child automaton are with its parent, then the child can be replaced using parameters.
Instead of binding the new child to the exact same actions of the parent that are being used by the old child, which is the root problem in the preceding scenario, the new child is bound to actions parameterized by the new child.
The parameters serve to differentiate the actions and obviate the need to wait for the old child to be destroyed.
To use this technique, the parent remembers the parameter associated with the current child and becomes unresponsive to actions not bearing that parameter.

>>>>

\section{Programming Model\label{programming_model}}

In this section, we show how I/O automata can be used to construct component based systems using ioa++.
Section~\ref{representation} describes how I/O automata are represented in the C++ language.
We then discuss the mechanisms available for dynamic composition in Section~\ref{dynamic_composition}.
Section~\ref{constellations} concludes with techniques for managing dynamic constellations of automata.

@@

As was mentioned in Section~\ref{design}, all user automata inherit from the common base class (ioa::automaton) which implements the system action interface.
\ifjournal
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

\paragraph*{Declaring and initializing state variables}
\fi
The state variables of an automaton are expressed as member variables of its corresponding class or class template.
\ifjournal
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Member variables are declared private, which is in keeping with the notion that the state of each automaton is independent, and are initialized using a constructor.
The constructor can also bootstrap the scheduler using the ioa::schedule function discussed below.
\ifjournal
For the asynch\_lcr\_automaton, the flags are both set initially false and the UID is initialized and added to the send queue.
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the ioa::schedule function discussed below.
\fi

\paragraph*{Declaring/defining actions}
Actions are member variables that model the \emph{action} 
\ifjournal
concept.
\else
concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
\fi
Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
These traits are used to guide the execution of the methods by the model and to supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

@@

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
Another common idiom is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
\ifjournal
For the asynch\_lcr\_automaton, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this common member function.
\fi

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

@@

In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the basic versions presented in the previous discussion.
This list is not exhaustive but rather serves as a distillation of our experiences so far working with ioa++.

\paragraph*{System action events}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

\paragraph*{Error reporting}
The event semantics of I/O automata provide a natural mechanism for reporting errors.
An automaton that encounters an error can use an output action to signal the error.
The automaton can continue processing if the error is transient or stop if the error is permanent.
In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.

@@

If all of the bindings of a child automaton are with its parent, then the child can be replaced using parameters.
Instead of binding the new child to the exact same actions of the parent that are being used by the old child, which is the key problem in the preceding scenario, the new child is bound to actions parameterized by the new child.
The parameters serve to differentiate the actions and obviate the need to wait for the old child to be destroyed.
To use this technique, the parent remembers the parameter associated with the current child and becomes unresponsive to actions not bearing that parameter.
